---
title: "proj3"
author: "Tuo Liu"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, nlme, arsenal, mice, lattice, DataExplorer, reshape2)
```

### Study Description

This project uses data derived from a phase III randomized controlled trial comparing Iofexidine versus placebo for opioid withdrawal symptoms. This was an 11-day longitudinal study with balanced study design but substantial amount of missing data due to dropouts.

The Modified Himmelsbach Opiate Withdrawal Scale (MHOWS) was chosen as the primary outcome since it is an objective assessment of the severity of opioid withdrawal signs. A weighted point value was assigned to each of the discontinuous and continuous signs for each subject. The final MHOWS was calculated as a linear combination of the weighted points of withdrawal signs.

The primary predictor of interests were time measured as days after randomization and treatment arm (1 if Iofexidine group; 0 placebo). The study objectives of this study was to assess the pattern of change in the mean response over time and whether the pattern of change in the mean response over time differs between treatment arms.

### Summary Statistics: baseline demographics
```{r}
# data read-in
demo <- read.csv("./data/base_end.csv")
withdraw <- read.csv("./data/withdraw.csv") %>% distinct(PATIENT, treat, day, .keep_all = T)
emesis <- read.csv("./data/emesis.csv")
```

```{r results='asis', message=FALSE}
# data prep
demo<- demo %>% mutate(
            treat=factor(treat),
            gender=factor(gender),
            smoke=factor(smoke),
            oral_nasal_smoke=factor(oral_nasal_smoke),
            IV = factor(IV)
            )

levels(demo$gender) <- c("Male", "Female")
levels(demo$smoke) <- c("No", "Yes")
levels(demo$oral_nasal_smoke) <- c("No", "Yes")
levels(demo$IV) <- c("No", "Yes")


# descriptive statistics for table 1
my_labels <- list(
  treat = "treatment arm",
  gender = "Sex (N, %)",
  age = "Age at randomization (Mean ± S.D.)",
  educ_yr = "Years of education completed (Mean ± S.D.)",
  smoke = "Currently smoking cigarettes (N, %)",
  oral_nasal_smoke = "oral/nasal/smoke",
  days_use_30d = "Days of opiate use in last 30 days (Mean ± S.D.)"
)
# # attr(data$SEX,'label')  <- 'Gender'
# # attr(data$PERIOD,'label')  <- 'Examination Cycle'
# # attr(data$CVD,'label')  <- 'CVD Status'

my_controls <- tableby.control(
  test = F,
  total = T,
  #numeric.test = "anova", cat.test = "chisq",
  numeric.stats = c("meansd"),
  cat.stats = c("countpct"),
  stats.labels = list(
    meansd = "Mean (SD)",
    countpct = "N (%)"
  )
)

table <- arsenal::tableby(treat ~  age + gender + educ_yr  + smoke + oral_nasal_smoke + IV + days_use_30d, data = demo, control = my_controls)

summary(table,
  labelTranslations = my_labels,
  title = "Baseline Demographics",
  pfootnote=TRUE,
  results="asis",
  digits=1
)
```



### MHOWS score calculation: discontinuous signs  
- Q1-Q4: 1 point if 2; 0 point otherwise
- Q5-Q6: 3 points if 2; 0 otherwise
- Q7: 5 points if 2; 0 otherwise
- Q8A-C: 3 points if any of A-C has 1/2; 0 otherwise
- emesis: 0 point if no episode; 5 point if 1 episode; 10 points if 2; 15 if 3 episodes

```{r}
# emsis
emesis[emesis==0]<- NA
emesis$eipsode <- rowSums(!is.na(emesis[,4:21]))

emesis$emesis <- ifelse(emesis$eipsode==0, 0, 
                        ifelse(emesis$eipsode==1,5, 
                               ifelse(emesis$eipsode==2, 10, 15)))
# withdraw
withdraw$pts_Q1 <- ifelse(withdraw$Q1==2, 1, 
                          ifelse(withdraw$Q1==1, 0, NA))
withdraw$pts_Q2 <- ifelse(withdraw$Q2==2, 1, 
                          ifelse(withdraw$Q2==1, 0, NA))
withdraw$pts_Q3 <- ifelse(withdraw$Q3==2, 1, 
                          ifelse(withdraw$Q3==1, 0, NA))
withdraw$pts_Q4 <- ifelse(withdraw$Q4==2, 1, 
                          ifelse(withdraw$Q4==1, 0, NA))
withdraw$pts_Q5 <- ifelse(withdraw$Q5==2, 3, 
                          ifelse(withdraw$Q5==1, 0, NA))
withdraw$pts_Q6 <- ifelse(withdraw$Q6==2, 3, 
                          ifelse(withdraw$Q6==1, 0, NA))
withdraw$pts_Q7 <- ifelse(withdraw$Q7==2, 5, 
                          ifelse(withdraw$Q7==1, 0, NA))
withdraw$pts_Q8 <- ifelse(withdraw$Q8A %in% c(1,2)|withdraw$Q8B %in% c(1,2)|withdraw$Q8C %in% c(1,2), 3, 0)
withdraw$pts <-  rowSums(withdraw[,23:35])

# combine emsis score
data <- dplyr::full_join(withdraw %>% select(1:3, 36), emesis %>% select(1:3, 23) %>% arrange(treat), by = c("PATIENT"="PATIENT", "treat"="treat", "day"="STDYDAY"))

# calculate MHOWS
data$MHOWS <- rowSums(data[,4:5])
```


### Missing data imputation
```{r}
plot_missing(data)
histogram(data$MHOWS)
```

### Longitudianl analysis
#### analysis with raw data
##### Sample means and sds
```{r message=FALSE}
# summarize data
data.sum <- data %>% group_by(treat,day) %>% 
  summarise(n=n(),means=mean(MHOWS, na.rm=T),stdev=sd(MHOWS, na.rm=T)) 

data.sum %>% 
  knitr::kable(digits = 2)


# plot mean profile over time
scaleFUN <- function(x) sprintf("%.0f", x)

# %>% filter(day != c(1,2))
ggplot(data.sum,aes(x=day,y=means,group=treat,colour=treat)) +
  geom_point(size=3) + 
  geom_line() + 
  scale_color_manual(name="Treatment",values=c("dodgerblue4","darkorange")) +
  scale_x_continuous(labels=scaleFUN, limits = c(1, 10), breaks = c(1:10)) + 
  scale_y_continuous(labels=scaleFUN, limits = c(0, 50), breaks = c(5,10,15,20,25,30,35,40,45,50)) +
  labs(x="Day",y="Mean MHOWS",title="Mean MHOWS Over Time") + 
  theme_classic()

ggsave("mean_response.png", width = 6, height = 3)
```

##### sample variance-covariance matrix 
```{r}
data_wide <- tidyr::spread(data %>% select(1:3,6), key = day, value = MHOWS) %>% select(1:2, 5:10)

# calculate sample variance-covariance matrix
cov(data_wide[,3:8], use = "complete.obs")
```


#### complete-case analysis
```{r}
# categorize day
data$dayN <- as.factor(data$day)

# subset day3-8
dat <- data %>% filter(day %in% c(3:8))


# fit maximal model with categorized dayN
mod1 <- gls(MHOWS~dayN*treat, 
            data=dat,
            corr=corSymm(form=~1|PATIENT),
            weights=varIdent(form=~1|dayN),
            method="REML",
            na.action = na.omit)
mod1 %>% summary()

# inspect variance-covariance matrix
getVarCov(mod1)
```

We saw significant treatment-time interaction effects for day4-day6, indicating the change pattern in mean MHOWS score over time was different between the treatment and placebo group.

#### analysis with imputed data


### Sensitivity analysis with respect to Missing Data



