---
title: "proj3"
author: "Tuo Liu"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, nlme, arsenal, mice, lattice, DataExplorer, reshape2,broom.mixed, ggResidpanel,gplots, gdata)
```

### Study Description

This project uses data derived from a phase III randomized controlled trial comparing Iofexidine versus placebo for opioid withdrawal symptoms. This was an 11-day longitudinal study with balanced study design but substantial amount of missing data due to dropouts.

The Modified Himmelsbach Opiate Withdrawal Scale (MHOWS) was chosen as the primary outcome since it is an objective assessment of the severity of opioid withdrawal signs. A weighted point value was assigned to each of the discontinuous and continuous signs for each subject. The final MHOWS was calculated as a linear combination of the weighted points of withdrawal signs.

The primary predictor of interests were time measured as days after randomization and treatment arm (1 if Iofexidine group; 0 placebo). The study objectives of this study was to assess the pattern of change in the mean response over time and whether the pattern of change in the mean response over time differs between treatment arms.

### Summary Statistics: baseline demographics
Two participants were removed as they dropped out since day 3.
```{r}
# data read-in
demo <- read.csv("./data/base_end.csv") %>% filter(!PATIENT %in% c(28482, 930174))
withdraw <- read.csv("./data/withdraw.csv") %>% distinct(PATIENT, treat, day, .keep_all = T) %>% filter(!PATIENT %in% c(28482, 930174))
emesis <- read.csv("./data/emesis.csv") %>% filter(!PATIENT %in% c(28482, 930174))
```

```{r results='asis', message=FALSE}
# data prep
demo<- demo %>% mutate(
            treat=factor(treat),
            gender=factor(gender),
            smoke=factor(smoke),
            oral_nasal_smoke=factor(oral_nasal_smoke),
            IV = factor(IV)
            )

levels(demo$gender) <- c("Male", "Female")
levels(demo$smoke) <- c("No", "Yes")
levels(demo$oral_nasal_smoke) <- c("No", "Yes")
levels(demo$IV) <- c("No", "Yes")


# descriptive statistics for table 1
my_labels <- list(
  treat = "treatment arm",
  gender = "Sex (N, %)",
  age = "Age at randomization (Mean ± S.D.)",
  educ_yr = "Years of education completed (Mean ± S.D.)",
  smoke = "Currently smoking cigarettes (N, %)",
  oral_nasal_smoke = "oral/nasal/smoke",
  days_use_30d = "Days of opiate use in last 30 days (Mean ± S.D.)"
)
# # attr(data$SEX,'label')  <- 'Gender'
# # attr(data$PERIOD,'label')  <- 'Examination Cycle'
# # attr(data$CVD,'label')  <- 'CVD Status'

my_controls <- tableby.control(
  test = F,
  total = T,
  #numeric.test = "anova", cat.test = "chisq",
  numeric.stats = c("meansd"),
  cat.stats = c("countpct"),
  stats.labels = list(
    meansd = "Mean (SD)",
    countpct = "N (%)"
  )
)

table <- arsenal::tableby(treat ~  age + gender + educ_yr  + smoke + oral_nasal_smoke + IV + days_use_30d, data = demo, control = my_controls)

summary(table,
  labelTranslations = my_labels,
  title = "Baseline Demographics",
  pfootnote=TRUE,
  results="asis",
  digits=1
)
```



### MHOWS score calculation: discontinuous signs  
- Q1-Q4: 1 point if 2; 0 point otherwise
- Q5-Q6: 3 points if 2; 0 otherwise
- Q7: 5 points if 2; 0 otherwise
- Q8A-C: 3 points if any of A-C has 1/2; 0 otherwise
- emesis: 0 point if no episode; 5 point if 1 episode; 10 points if 2; 15 if 3 episodes

```{r}
# emsis
emesis[emesis==0]<- NA
emesis$eipsode <- rowSums(!is.na(emesis[,4:21]))

emesis$emesis <- ifelse(emesis$eipsode==0, 0, 
                        ifelse(emesis$eipsode==1,5, 
                               ifelse(emesis$eipsode==2, 10, 15)))
# withdraw
withdraw$pts_Q1 <- ifelse(withdraw$Q1==2, 1, 
                          ifelse(withdraw$Q1==1, 0, NA))
withdraw$pts_Q2 <- ifelse(withdraw$Q2==2, 1, 
                          ifelse(withdraw$Q2==1, 0, NA))
withdraw$pts_Q3 <- ifelse(withdraw$Q3==2, 1, 
                          ifelse(withdraw$Q3==1, 0, NA))
withdraw$pts_Q4 <- ifelse(withdraw$Q4==2, 1, 
                          ifelse(withdraw$Q4==1, 0, NA))
withdraw$pts_Q5 <- ifelse(withdraw$Q5==2, 3, 
                          ifelse(withdraw$Q5==1, 0, NA))
withdraw$pts_Q6 <- ifelse(withdraw$Q6==2, 3, 
                          ifelse(withdraw$Q6==1, 0, NA))
withdraw$pts_Q7 <- ifelse(withdraw$Q7==2, 5, 
                          ifelse(withdraw$Q7==1, 0, NA))
withdraw$pts_Q8 <- ifelse(withdraw$Q8A %in% c(1,2)|withdraw$Q8B %in% c(1,2)|withdraw$Q8C %in% c(1,2), 3, 0)
withdraw$pts <-  rowSums(withdraw[,23:35])

# combine emsis score
data <- dplyr::full_join(withdraw %>% select(1:3, 36), emesis %>% select(1:3, 23) %>% arrange(treat), by = c("PATIENT"="PATIENT", "treat"="treat", "day"="STDYDAY"))

# calculate MHOWS
data$MHOWS <- rowSums(data[,4:5])
```



### Longitudianl analysis
#### Sample means and sds
```{r message=FALSE}
# summarize data
data.sum <- data %>% group_by(treat,day) %>% 
  summarise(n=n(),means=mean(MHOWS, na.rm=T),stdev=sd(MHOWS, na.rm=T)) 

data.sum %>% 
  knitr::kable(digits = 2)


# plot mean profile over time
scaleFUN <- function(x) sprintf("%.0f", x)

# %>% filter(day != c(1,2))
ggplot(data.sum %>% filter(!day %in% c(1,2, 9,10 )),aes(x=day ,y=means,group=treat,colour=treat)) +
  geom_point(size=3) +
  geom_line() +
  scale_color_manual(name="Treatment",values=c("dodgerblue4","darkorange")) +
  scale_x_continuous(labels=scaleFUN, limits = c(1, 10), breaks = c(1:10)) +
  scale_y_continuous(labels=scaleFUN, limits = c(0, 50), breaks = c(5,10,15,20,25,30,35,40,45,50)) +
  labs(x="Day",y="Mean MHOWS",title="Mean MHOWS Over Time") +
  theme_classic()

ggsave("mean_response.png", width = 6, height = 3)
```

#### Primary analysis with imputed data
##### sample variance-covariance matrix 
Some things are easier in the wide format, e.g., change scores or imputing data, while other procedures are easier in the long format, e.g., graphics and advanced statistical modeling. It is therefore useful to have both formats available.
```{r}
data_wide <- tidyr::spread(data %>% select(1:3,6), key = day, value = MHOWS) %>% select(1:2, 5:10) %>% mutate(treat=factor(treat))

# calculate sample variance-covariance matrix
cov(data_wide[,3:8], use = "complete.obs") %>% knitr::kable(digits = 2)
```

##### Missing pattern
```{r}
png(filename="./missing_pattern.png",  width=480, height=480)
md.pattern(data_wide, rotate.names = T)
dev.off()
```
The missingness pattern shows that there are 170 missing values in total: 5 for day4 , 30 for day5, 40 for day6, 47 for day7,  and 48 for day8. Moreover, there are 16 completely observations. The overall missing pattern was non-monotone with 48 cases followed a "monotone" missing pattern and 2 followed "non-monotone pattern". 




##### Mixed-MI: categorical time comparing post-treat timepoint to baseline
```{r}
colnames(data_wide) <- c("PATIENT","treat",paste0("day", 3:8))

# Create imputations, using predictorMatrix to specify the
# variables used in imputations
ini <- mice(data_wide,
            maxit = 0)
pred <- ini$predictorMatrix
pred[,"PATIENT"] <- 0

# impute data in wide format
imp_dat <- mice(data_wide,
                printFlag = FALSE,
                predictorMatrix = pred,
                seed = 2022)

# double-check imputation method used for repeated measurements
# imp_dat$method

# summary meansd
# calculate some descriptive statistics
imp_dat_wide <- complete(imp_dat, include = F, action = "long")

imp_sum <- imp_dat_wide %>% 
  melt(
    id.vars=c(".imp", "PATIENT", "treat"), value.name="y", measure.vars=c("day3", "day4", "day5","day6","day7", "day8")) %>%
  group_by(.imp, treat, variable) %>%
  summarize(n=n(),
            mean=mean(y),
            sd=sd(y))
levels(imp_sum$variable) <- 3:8
imp_sum$day<- as.numeric(imp_sum$variable)+2
# plot
ggplot(imp_sum,aes(x=day,y=mean,group=treat,colour=treat)) +
  geom_point(size=3) +
  geom_line() +
  scale_color_manual(name="Treatment",values=c("dodgerblue4","darkorange")) +
  scale_x_continuous(limits = c(1, 10), breaks = c(1:10)) +
  scale_y_continuous(labels=scaleFUN, limits = c(0, 50), breaks = c(5,10,15,20,25,30,35,40,45,50)) +
  labs(x="Day",y="Mean MHOWS",title="Mean MHOWS Over Time") +
  theme_classic()+
  facet_wrap(~.imp, nrow=5)

ggsave("mean_response_imp.png", width = 6, height = 8)


# fit the models to the imputed data
mod11 <- with(imp_dat,
             {
               dat <- data.frame(treat = treat,
                                 PATIENT = PATIENT,
                                 day3 = day3,
                                 day4 = day4,
                                 day5 = day5,
                                 day6 = day6,
                                 day7 = day7,
                                 day8 = day8)
               dat_long <- pivot_longer(dat,
                                        cols = contains("day"),
                                        names_to = "day",
                                        values_to = "y")
               lme(y ~ day*treat,
                   random = ~1|PATIENT,
                   data = dat_long)
             })

mod11_df <- summary(mice::pool(mod11),conf.int = TRUE) 
mod11_df %>% knitr::kable(digits = 2)
```
##### Mixed-MI: numerical time-parametric trend
```{r}
# fit the models to the imputed data
mod22 <- with(imp_dat,
             {
               dat <- data.frame(treat = treat,
                                 PATIENT = PATIENT,
                                 `3` = day3,
                                 `4` = day4,
                                 `5` = day5,
                                 `6` = day6,
                                 `7` = day7,
                                 `8` = day8,
                                 check.names=FALSE)
               
               dat_long <- melt(dat,
                                id.vars = c("PATIENT", "treat"),
                                value.name = "y") %>% mutate(day=ifelse(variable==3, 0,
                                                                        ifelse(variable==4, 1, 
                                                                               ifelse(variable==5, 2, 
                                                                                      ifelse(variable==6,3,
                                                                                             ifelse(variable==7,4,5))))))

               lme(y ~ day*treat,
                   random = ~1|PATIENT,
                   data = dat_long)
             })

mod22_df <- summary(mice::pool(mod22),conf.int = TRUE) 
mod22_df %>% knitr::kable(digits = 2)
```


### Sensitivity analysis with respect to Missing Data
#### complete-case analysis
```{r}
# categorize day
data$dayN <- as.factor(data$day)

# subset day3-8
dat <- data %>% filter(day %in% c(3:8))


# fit maximal model with categorized dayN
mod1 <- lme(MHOWS ~ dayN*treat,
            random = ~1|PATIENT,
            data = dat,
            na.action = na.omit)

mod1_df <- as.data.frame(coef(summary(mod1)))
mod1_df %>% knitr::kable(digits = 2)
# inspect variance-covariance matrix
# getVarCov(mod1)  %>% knitr::kable(digits = 2)
```
We saw significant treatment-time interactions on day4-day6, indicating the change pattern in mean MHOWS score over time was different between the treatment and placebo group.

```{r}
# continuous day
# subset day3-8
dat <- data %>% filter(day %in% c(3:8))


# fit maximal model with categorized dayN
mod2 <- lme(MHOWS ~ day*treat,
            random = ~1|PATIENT,
            data = dat,
            na.action = na.omit)
mod2_df <- as.data.frame(coef(summary(mod2)))
mod2_df %>% knitr::kable(digits = 2)
# inspect variance-covariance matrix
# getVarCov(mod2) %>% knitr::kable(digits = 2)
```

#### effect comparison between analysis after MICE imputation and Complete-case analysis
```{r}
c2 <- mod2_df %>% mutate(term=row.names(.)) %>% select(6,1,2,5) %>% rename(estimate=Value, std.error=Std.Error, p.value=`p-value`)
mi2 <- mod22_df %>% select(1:3,6)
df2 <- gdata::combine(mi2, c2)
df2$term <- str_replace(df2$term, "treatPLACEBO", "Placebo")


c1 <- mod1_df %>% mutate(term=row.names(.)) %>% select(6,1,2,5) %>% rename(estimate=Value, std.error=Std.Error, p.value=`p-value`)
mi1 <- mod11_df %>% select(1:3,6)
df1 <- gdata::combine(mi1, c1)

df1$term <- str_replace(df1$term, "dayday", "dayN")
df1$term <- str_replace(df1$term, "treatPLACEBO", "Placebo")


df <- gdata::combine(df1,df2)
df$source.1 <- factor(df$source.1, levels = c("df1", "df2"),
                  labels = c("Categorical time", "Linear Trend")
                  )
# plot
pd <- position_dodge(0.1) # move them .05 to the left and right

ggplot(df, aes(x=term, y=estimate, colour=source, group=source)) + 
    geom_errorbar(aes(ymin=estimate-std.error, ymax=estimate+std.error), colour="black", width=.1, position=pd) +
    geom_point(position=pd, size=3, shape=21, fill="white") + # 21 is filled circle
  facet_wrap(~source.1, scales = "free") +
    xlab("Term") +
    ylab("Estimate") +
    scale_colour_hue(name="Analysis choice",    # Legend label, use darker colors
                     labels=c("Complete-case", "MICE"),
                     breaks=c("mi1", "c1"),
                     l=40) +                    # Use darker colors, lightness=40
    expand_limits(y=0) +                        # Expand y range
    scale_y_continuous(breaks=0:20*4) +         # Set tick every 4
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          legend.position=c(.8,.8)) 


# ggsave("est comparison.png", width = 8, height = 4)
```


#### effect comparison for m
categorical time model
```{r}
imp_dat_10 <- mice(data_wide, m=10,
                printFlag = FALSE,
                predictorMatrix = pred,
                seed = 2022)
imp_dat_20 <- mice(data_wide, m=20,
                printFlag = FALSE,
                predictorMatrix = pred,
                seed = 2022)
imp_dat_50 <- mice(data_wide, m=50,
                printFlag = FALSE,
                predictorMatrix = pred,
                seed = 2022)

# fit the models to the imputed data
# m=10
mod110 <- with(imp_dat_10,
             {
               dat <- data.frame(treat = treat,
                                 PATIENT = PATIENT,
                                 day3 = day3,
                                 day4 = day4,
                                 day5 = day5,
                                 day6 = day6,
                                 day7 = day7,
                                 day8 = day8)
               dat_long <- pivot_longer(dat,
                                        cols = contains("day"),
                                        names_to = "day",
                                        values_to = "y")
               lme(y ~ day*treat,
                   random = ~1|PATIENT,
                   data = dat_long)
             })

mod110_df <- summary(mice::pool(mod110),conf.int = TRUE) 
mod110_df %>% knitr::kable(digits = 2)

# m=20
mod120 <- with(imp_dat_20,
             {
               dat <- data.frame(treat = treat,
                                 PATIENT = PATIENT,
                                 day3 = day3,
                                 day4 = day4,
                                 day5 = day5,
                                 day6 = day6,
                                 day7 = day7,
                                 day8 = day8)
               dat_long <- pivot_longer(dat,
                                        cols = contains("day"),
                                        names_to = "day",
                                        values_to = "y")
               lme(y ~ day*treat,
                   random = ~1|PATIENT,
                   data = dat_long)
             })

mod120_df <- summary(mice::pool(mod120),conf.int = TRUE) 
mod120_df %>% knitr::kable(digits = 2)

# m=50
mod150 <- with(imp_dat_50,
             {
               dat <- data.frame(treat = treat,
                                 PATIENT = PATIENT,
                                 day3 = day3,
                                 day4 = day4,
                                 day5 = day5,
                                 day6 = day6,
                                 day7 = day7,
                                 day8 = day8)
               dat_long <- pivot_longer(dat,
                                        cols = contains("day"),
                                        names_to = "day",
                                        values_to = "y")
               lme(y ~ day*treat,
                   random = ~1|PATIENT,
                   data = dat_long)
             })

mod150_df <- summary(mice::pool(mod150),conf.int = TRUE) 
mod150_df %>% knitr::kable(digits = 2)
```
linear trend model
```{r}
# imp_dat_10 <- mice(data_wide, m=10,
#                 printFlag = FALSE,
#                 predictorMatrix = pred,
#                 seed = 2022)
# imp_dat_20 <- mice(data_wide, m=20,
#                 printFlag = FALSE,
#                 predictorMatrix = pred,
#                 seed = 2022)
# imp_dat_50 <- mice(data_wide, m=50,
#                 printFlag = FALSE,
#                 predictorMatrix = pred,
#                 seed = 2022)

# fit the models to the imputed data
# m=10
mod210 <- with(imp_dat_10,
             {
               dat <- data.frame(treat = treat,
                                 PATIENT = PATIENT,
                                 `3` = day3,
                                 `4` = day4,
                                 `5` = day5,
                                 `6` = day6,
                                 `7` = day7,
                                 `8` = day8,
                                 check.names=FALSE)
               
               dat_long <- melt(dat,
                                id.vars = c("PATIENT", "treat"),
                                value.name = "y") %>% mutate(day=ifelse(variable==3, 0,
                                                                        ifelse(variable==4, 1, 
                                                                               ifelse(variable==5, 2, 
                                                                                      ifelse(variable==6,3,
                                                                                             ifelse(variable==7,4,5))))))

               lme(y ~ day*treat,
                   random = ~1|PATIENT,
                   data = dat_long)
             })

mod210_df <- summary(mice::pool(mod210),conf.int = TRUE) 
mod210_df %>% knitr::kable(digits = 2)

# m=20
mod220 <- with(imp_dat_20,
             {
               dat <- data.frame(treat = treat,
                                 PATIENT = PATIENT,
                                 `3` = day3,
                                 `4` = day4,
                                 `5` = day5,
                                 `6` = day6,
                                 `7` = day7,
                                 `8` = day8,
                                 check.names=FALSE)
               
               dat_long <- melt(dat,
                                id.vars = c("PATIENT", "treat"),
                                value.name = "y") %>% mutate(day=ifelse(variable==3, 0,
                                                                        ifelse(variable==4, 1, 
                                                                               ifelse(variable==5, 2, 
                                                                                      ifelse(variable==6,3,
                                                                                             ifelse(variable==7,4,5))))))

               lme(y ~ day*treat,
                   random = ~1|PATIENT,
                   data = dat_long)
             })

mod220_df <- summary(mice::pool(mod220),conf.int = TRUE) 
mod220_df %>% knitr::kable(digits = 2)

# m=50
mod250 <- with(imp_dat_50,
             {
               dat <- data.frame(treat = treat,
                                 PATIENT = PATIENT,
                                 `3` = day3,
                                 `4` = day4,
                                 `5` = day5,
                                 `6` = day6,
                                 `7` = day7,
                                 `8` = day8,
                                 check.names=FALSE)
               
               dat_long <- melt(dat,
                                id.vars = c("PATIENT", "treat"),
                                value.name = "y") %>% mutate(day=ifelse(variable==3, 0,
                                                                        ifelse(variable==4, 1, 
                                                                               ifelse(variable==5, 2, 
                                                                                      ifelse(variable==6,3,
                                                                                             ifelse(variable==7,4,5))))))

               lme(y ~ day*treat,
                   random = ~1|PATIENT,
                   data = dat_long)
             })

mod250_df <- summary(mice::pool(mod250),conf.int = TRUE) 
mod250_df %>% knitr::kable(digits = 2)
```
```{r}
# stack mod_df
df_1 <- gdata::combine(mod11_df,mod110_df, mod120_df, mod150_df)
df_1$source <- factor(df_1$source)
levels(df_1$source) = c("m=5", "m=10", "m=20", "m=50")
                  
df_1$term <- str_replace(df_1$term, "dayday", "dayN")
df_1$term <- str_replace(df_1$term, "treatPLACEBO", "Placebo")                  


df_2 <- gdata::combine(mod22_df,mod210_df, mod220_df, mod250_df)
df_2$source <- factor(df_2$source)
levels(df_2$source) = c("m=5", "m=10", "m=20", "m=50")


df_2$term <- str_replace(df_2$term, "dayday", "dayN")
df_2$term <- str_replace(df_2$term, "treatPLACEBO", "Placebo")


# plot
pd <- position_dodge(0.5) # move them .05 to the left and right

ggplot(df_1, aes(x=term, y=estimate, colour=source, group=source)) + 
    geom_errorbar(aes(ymin=estimate-std.error, ymax=estimate+std.error), colour="black", width=.1, position=pd) +
    geom_point(position=pd, size=3, shape=21, fill="white") + # 21 is filled circle
    xlab("Term") +
    ylab("Estimate") +
    scale_colour_hue(name="Number of Imputation",    # Legend label, use darker colors
                     labels=c("5", "10", "20", "50"),
                     l=40) +                    # Use darker colors, lightness=40
    expand_limits(y=0) +                        # Expand y range
    scale_y_continuous(breaks=0:20*4) +         # Set tick every 4
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          legend.position="top", legend.box = "horizontal") 


ggsave("est_comparison_1.png", width = 8, height = 4)

ggplot(df_2, aes(x=term, y=estimate, colour=source, group=source)) + 
    geom_errorbar(aes(ymin=estimate-std.error, ymax=estimate+std.error), colour="black", width=.1, position=pd) +
    geom_point(position=pd, size=3, shape=21, fill="white") + # 21 is filled circle
    xlab("Term") +
    ylab("Estimate") +
    scale_colour_hue(name="Number of Imputation",    # Legend label, use darker colors
                     labels=c("5", "10", "20", "50"),
                     l=40) +                    # Use darker colors, lightness=40
    expand_limits(y=0) +                        # Expand y range
    scale_y_continuous(breaks=0:20*4) +         # Set tick every 4
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          legend.position="top", legend.box = "horizontal") 

ggsave("est_comparison_2.png", width = 8, height = 4)
```





### diagnostics
```{r}
# checking on distribution of response
png(filename="./MHOWS_his.png",  width=600)
histogram(data$MHOWS)
dev.off()
```

#### mixed model diagnostics
```{r}
# select the first complete dataset to fit lme
imp_dat_wide_1 <- imp_dat_wide %>% filter(.imp==1) %>% 
  melt(id.vars=c("PATIENT", "treat"), value.name="y", 
       measure.vars=c("day3", "day4", "day5","day6","day7", "day8"))
levels(imp_dat_wide_1$variable) <- 3:8
imp_dat_wide_1$day<- as.numeric(imp_dat_wide_1$variable)+2


mod_s <- lme(y ~ day*treat,
            random = ~1|PATIENT,
            data = imp_dat_wide_1)

png(filename="./resid_hist.png",  width=600)
resid_panel(mod_s, plots=c("hist"), type = "response")
dev.off()

png(filename="./qq.png",  width=600)
resid_panel(mod_s, plots=c("qq"), type = "response")
dev.off()

png(filename="./resid-fitted.png",  width=600)
resid_panel(mod_s, plots=c("resid"),  smoother=T, type = "response")
dev.off()
```


