---
title: "code"
author: "Tuo Liu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
## load packages from CRAN
pacman::p_load(rio,          # File import
               here,         # File locator
               tidyverse,    # data management + ggplot2 graphics
               tsibble,      # handle time series datasets
               survey,       # for survey functions
               srvyr,        # dplyr wrapper for survey package
               gtsummary,    # wrapper for survey package to produce tables
               apyramid,     # a package dedicated to creating age pyramids
               patchwork,    # for combining ggplots
               ggforce,      # for alluvial/sankey plots
               survival,
               janitor
               ) 

## load packages from github
pacman::p_load_gh(
     "R4EPI/sitrep"          # for observation time / weighting functions
)
```

## Introduction

Survival analysis focuses on describing a defined point of event called the failure (onset of a disease, cure from a disease, death...) that occurs after a period of time called failure time during which individuals are observed, for a given individual or group of individuals.

A time of origin or the start of observation is necessary to determine the failure time. The inferential target for survial analysis is the time between an origin and an event.  For example, in clinical studies it is used to assess the effect of a treatment for instance, or to assess a large variety of cancer survival measures.  

Censoring: Censoring occurs when at the end of follow-up, some of the individuals have not had the event of interest, and thus their true time to event is unknown.

## Data Prep
```{r}
# import linelist
linelist_case_data <- rio::import("linelist_cleaned.rds")
```

survival data can be described as having the following three characteristics:

- the dependent variable or response is the waiting time until the occurrence of a well-defined event,
- observations are censored, in the sense that for some units the event of interest has not occurred at the time the data are analyzed, and
- there are predictors or explanatory variables whose effect on the waiting time we wish to assess or control.

```{r}
#create a new data called linelist_surv from the linelist_case_data

linelist_surv <-  linelist_case_data %>% 
  dplyr::filter(
       # remove observations with wrong or missing dates of onset or date of outcome
       date_outcome > date_onset) %>% 
  dplyr::mutate(
       # create the event var which is 1 if the patient died and 0 if he was right censored
       event = ifelse(is.na(outcome) | outcome == "Recover", 0, 1),
       # create the var on the follow-up time in days
       futime = as.double(date_outcome - date_onset),
       # create a new age category variable with only 3 strata levels
       age_cat_small = dplyr::case_when( 
            age_years < 5  ~ "0-4",
            age_years >= 5 & age_years < 20 ~ "5-19",
            age_years >= 20   ~ "20+"),
       # previous step created age_cat_small var as character.
       # now convert it to factor and specify the levels.
       # Note that the NA values remain NA's and are not put in a level "unknown" for example,
       # since in the next analyses they have to be removed.
       age_cat_small = fct_relevel(age_cat_small, "0-4", "5-19", "20+")
       )

summary(linelist_surv$futime)
linelist_surv %>% tabyl(outcome, event)
linelist_surv %>% tabyl(age_cat_small, age_cat)
linelist_surv %>% 
  select(case_id, age_cat_small, date_onset, date_outcome, outcome, event, futime) %>% 
  head(10)
```

## Survival Analysis Basics

- Building a surv-type object

```{r}
# Use Suv() syntax for right-censored data
survobj <- Surv(time = linelist_surv$futime,
                event = linelist_surv$event)
head(survobj, 10)
```
- running initial analyses
  - survfit object
  - life table
    - the number of people who were at risk of developing the event (people who did not
    have the event yet nor were censored: n.risk)
    - those who did develop the event (n.event)
    - and from the above: the probability of not developing the event (probability of not
    dying, or of surviving past that specific time)
    - the standard error and the confidence interval for that probability are derived and
    displayed

```{r}
# fit the KM estimates using a formula where the Surv object "survobj" is the response variable.
# "~ 1" signifies that we run the model for the overall survival  
linelistsurv_fit <-  survival::survfit(survobj ~ 1)

#print its summary at specific times
summary(linelistsurv_fit, times = c(5,10,20,30,60))
```
- cumulative hazard: the instantaneous rate of event occurrence
- plotting Kaplan-Meir curves

```{r}
# original plot
plot(
  linelistsurv_fit,
  xlab = "Days of follow-up",       
  ylab = "Survival Probability",       
  mark.time = TRUE,              # mark events on the curve: a "+" is printed at every event
  conf.int = FALSE,              # do not plot the confidence interval
  main = "Overall survival curve and cumulative mortality"
  )

# draw an additional curve to the previous plot
lines(
  linelistsurv_fit,
  lty = 3,             # use different line type for clarity
  fun = "event",       # draw the cumulative events instead of the survival 
  mark.time = FALSE,
  conf.int = FALSE
  )

# add a legend to the plot
legend(
  "topright",                               # position of legend
  legend = c("Survival", "Cum. Mortality"), # legend text 
  lty = c(1, 3),                            # line types to use in the legend
  cex = .85,                                # parametes that defines size of legend text
  bty = "n"                                 # no box type to be drawn for the legend
  )
```


## Comparison of survival curves

To compare the survival within different groups of our observed participants or patients, we might need to first look at their respective survival curves and then run tests to evaluate the difference between independent groups.


- Log rank test

```{r}
# create the new survfit object based on gender
linelistsurv_fit_sex <-  survfit(Surv(futime, event) ~ gender, data = linelist_surv)

# set colors
col_sex <- c("lightgreen", "red")

# create plot
plot(
  linelistsurv_fit_sex,
  col = col_sex,
  xlab = "Days of follow-up",
  ylab = "Survival Probability")

# add legend
legend(
  "topright",
  legend = c("Female","Male"),
  col = col_sex,
  lty = 1,
  cex = .9,
  bty = "n")

#compute the test of the difference between the survival curves
survival::survdiff(
  Surv(futime, event) ~ gender, 
  data = linelist_surv
  )
```
Some other R packages allow illustrating survival curves for different groups and testing the difference all at once.

```{r warning=FALSE}

survminer::ggsurvplot(
    linelistsurv_fit_sex, 
    data = linelist_surv,          # again specify the data used to fit linelistsurv_fit_sex 
    conf.int = FALSE,              # do not show confidence interval of KM estimates
    surv.scale = "percent",        # present probabilities in the y axis in %
    break.time.by = 10,            # present the time axis with an increment of 10 days
    xlab = "Follow-up days",
    ylab = "Survival Probability",
    pval = T,                      # print p-value of Log-rank test 
    pval.coord = c(40,.91),        # print p-value at these plot coordinates
    risk.table = T,                # print the risk table at bottom 
    legend.title = "Gender",       # legend characteristics
    legend.labs = c("Female","Male"),
    font.legend = 10, 
    palette = "Dark2",             # specify color palette 
    surv.median.line = "hv",       # draw horizontal and vertical lines to the median survivals
    ggtheme = theme_light()        # simplify plot background
)
```

```{r warning=FALSE}
linelistsurv_fit_source <-  survfit(
  Surv(futime, event) ~ source,
  data = linelist_surv
  )

# plot
survminer::ggsurvplot( 
  linelistsurv_fit_source,
  data = linelist_surv,
  size = 1, linetype = "strata",   # line types
  conf.int = T,
  surv.scale = "percent",  
  break.time.by = 10, 
  xlab = "Follow-up days",
  ylab= "Survival Probability",
  pval = T,
  pval.coord = c(40,.91),
  risk.table = T,
  legend.title = "Source of \ninfection",
  legend.labs = c("Funeral", "Other"),
  font.legend = 10,
  palette = c("#E7B800","#3E606F"),
  surv.median.line = "hv", 
  ggtheme = theme_light()
)
```

## Cox regression analysis

Cox proportional hazards regression is one of the most popular regression techniques for survival analysis. Other models can also be used since the Cox model requires important assumptions that need to be verified for an appropriate use. 

In a Cox proportional hazards regression model, the measure of effect is the hazard rate (HR), which is the risk of failure (or the risk of death in our example), given that the participant has survived up to a specific time. Usually, we are interested in comparing independent groups with respect to their hazards, and we use a hazard ratio, which is analogous to an odds ratio in the setting of multiple logistic regression analysis. 

```{r}
#fitting the cox model
linelistsurv_cox_sexage <-  survival::coxph(
              Surv(futime, event) ~ gender + age_cat_small, 
              data = linelist_surv
              )


#summary of the model
summary(linelistsurv_cox_sexage)
```
The function `cox.zph()` from survival package may be used to test the proportional hazards assumption for a Cox regression model fit.

```{r}
test_ph_sexage <- survival::cox.zph(linelistsurv_cox_sexage)
test_ph_sexage
```
```{r}
#fit the model
linelistsurv_cox <-  coxph(
                        Surv(futime, event) ~ gender + age_years + 
                          source + days_onset_hosp,
                        data = linelist_surv
                        )


#test the proportional hazard model
linelistsurv_ph_test <- cox.zph(linelistsurv_cox)
linelistsurv_ph_test
```
The graphical verification of this assumption may be performed with the function ggcoxzph() from the survminer package.

```{r}
survminer::ggcoxzph(linelistsurv_ph_test)
```

```{r}
survminer::ggforest(linelistsurv_cox, data = linelist_surv)
```

