---
title: "descriptive statistics"
author: "Tuo Liu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, DataExplorer, arsenal, matchIt)
```

### study description
#### 1.	brief discussion of the objectives of the project and the sample

Objective: to investigate the association between smoking cessation and weight gain.   
The sample is relative large (N=1629), including information from interviews, examinations. 63 subjects were excluded due to missing value of weight difference between 1971 and 1982, and extra 41 subjects were removed because of weight loss medication. Table 1 presented a summary of the rest of samples (N=1525) by stratification on weight gain status (Yes/No). 

#### 2. identify and define outcomes, the exposure or treatment and important predictors

The major predictor was smoking cessation (1 if subject quitted smoking during study, 0 otherwise). Other important predictors included subjects’ gender, age, race/ethnicity, BMI, alcohol consumption.  

The outcome variable was subjects' body weight. 

### data read-in
```{r}
data <- read.csv("./data/nhefs.csv") %>% filter(!is.na(wt82_71) & wtloss != 1)
colnames(data)
```


### descriptive statistics
```{r results='asis'}
# create weight gain status variable
data$wtgain <- ifelse(data$wt82_71>0, "Yes", "No")

# data prep
data <- data %>% mutate(
            qsmk=factor(qsmk),
            sex=factor(sex),
            race=factor(race),
            education=factor(education)
            )

levels(data$qsmk) <- c("No", "Yes")
levels(data$sex) <- c("Male", "Female")
levels(data$race) <- c("White", "Other")
levels(data$education) <- c("8th grade or less", "High school DROPOUT", "High school","College dropout", "College or more")


# descriptive statistics for table 1
my_labels <- list(
  qsmk = "Quit smoking between 1971 and 1982",
  wtgain = "Gained weight from 1971",
  sex = "Gender",
  age = "Age (yrs) in 1971",
  race = "Race",
  education = "Education by 1971",
  smkintensity82_71 = "Increase in number of cigarettes/day from 1971 to 1982",
  smokeyrs = "YEARS OF SMOKING"
)
# # attr(data$SEX,'label')  <- 'Gender'
# # attr(data$PERIOD,'label')  <- 'Examination Cycle'
# # attr(data$CVD,'label')  <- 'CVD Status'
# 
# 
my_controls <- tableby.control(
  test = F,
  total = F,
  #numeric.test = "anova", cat.test = "chisq",
  numeric.stats = c("meansd"),
  cat.stats = c("countpct"),
  stats.labels = list(
    meansd = "Mean (SD)"
  )
)

table <- arsenal::tableby(wtgain ~  sex + age + race + education + qsmk + smokeyrs + smkintensity82_71, data = data, control = my_controls)

summary(table,
  labelTranslations = my_labels,
  title = "Summary Statistics of NHEFS 1971-1982 Data",
  pfootnote=TRUE,
  results="asis",
  digits=1
)
```


### descriptive statistic_Yujung
```{r}
# table(data$qsmk)
# n_miss(data$qsmk)

data$status <- as.factor(ifelse(data$qsmk == 0 & data$smokeintensity <=10, 'noqui-light',
                                ifelse(data$qsmk == 0 & data$smokeintensity <20, 'noqui-moderate',
                                       ifelse(data$qsmk == 0 & data$smokeintensity >= 20, 'noqui-heavy',
                                              ifelse(data$qsmk ==1 & data$smokeintensity <=10, 'qui-light',
                                                     ifelse(data$qsmk == 1 & data$smokeintensity <20, 'qui-moderate',
                                                            ifelse(data$qsmk == 1 & data$smokeintensity >=20, 'qui-heavy', 'other')))))))


data$chol_status <- as.factor(ifelse(data$cholesterol <200, 'Less than 200 mg/dL', 'At or above 200 mg/dL'))

data$sex <- as.factor(data$sex)
levels(data$sex) <- c("Male", "Female")
data$hbp <- as.factor(data$hbp)
levels(data$hbp) <- c("No", "Yes", "Missing" )
data$diabetes <- as.factor(data$diabetes)
levels(data$diabetes) <- c("No", "Yes", "Missing" )
data$wtloss <- as.factor(data$wtloss)
levels(data$wtloss) <- c("No", "Yes")



# table(data$status)
# n_miss(data$status)

# Create table
controls <- tableby.control( test = TRUE, total = TRUE,
                             numeric.test="kwt", cat.test="chisq",
                             numeric.stats = c("N", "meansd", "Nmiss"),
                             cat.stats = c("countpct" , "Nmiss"),
                             stats.labels=list(N='Count', Nmiss='Missing'))

Mylabels <- list(chol_status = "Cholesterol level", wt82_71 = "Weight change", age = "Age", sex = "Gender", hbp = "Dx Hight blood pressure", diabetes = "Dx Diabetes", wtloss = "Used weight loss medication" )

table2 <- tableby(status ~ wt82_71 + age + sex + chol_status + hbp + diabetes + wtloss, data = data, control = controls )


```

```{r results = "asis"}
summary(table2, text=TRUE, labelTranslations = Mylabels)
```


### Propensity Score 
1. Definition:
- A propensity score is a single score that represents the probability of receiving a treatment, conditional on a set of observed covariates.  

2. Objective
- The goal of creating a propensity score is to balance covariates between individuals who did and did not receive a treatment, making it easier to isolate the effect of a treatment.

- Propensity scores are used to reduce confounding and thus include variables thought to be related to both treatment and outcome. 

3. Use case
- Propensity scores are useful when estimating a treatment’s effect on an outcome using observational data and when selection bias due to nonrandom treatment assignment is likely.

- RCT:  control individuals that are similar to treated individuals in everything but treatment receipt are used as proxies for the counterfactual.

- In observational data, however, treatment assignment is not random. This leads to selection bias, where measured and unmeasured characteristics of individuals are associated with likelihood
of receiving treatment and with the outcome. 

4. Example
- We used propensity scores to account for the fact that patients’ baseline health affects both probability of receiving palliative care and experiencing adverse health outcomes.(see example in help file)

- if one were concerned that age might affect both treatment selection and outcome, one strategy would be to compare individuals of similar age in both treatment and comparison groups.

- But it's unlikely to find individuals with identical age, gender, race, insurrance status, etc from treatmetn and control group.

- Compress these factors into one single score

5. STEPS INVOLVED IN CONSTRUCTING AND ASSESSING PROPENSITY SCORES
- Step One: Choice of Variables to Include in the Propensity Score
  - Guidance: controlling for covariates associated with the outcome
  - If a variable is thought to be related to the __outcome__ but not the treatment, including it in the propensity score should reduce bias.
    - With sufficiently large datasets, it is beneficial to include all variables that are potentially related to the outcome.
    - In smaller datasets, however, potentially irrelevant covariates may introduce too much “noise” into treatment effect estimates
  - Controlling for variables that are hypothesized to be associated with _treatment_ but _not outcome_, however, can __decrease precision__.
  
  
- Step Two: Balance of Propensity Score across Treatment and Comparison Groups
  
  
- Step Three: Balance of Covariates across Treatment and Comparison Groups within Blocks of the Propensity Score
  
  
- Step Four: Choice of Matching and Weighting Strategies


- Step Five: Balance of Covariates after Matching or Weighting the Sample by a Propensity Score
  
  
  
### Example: [the effect of going to Catholic school, as opposed to public school, on student achievement](https://sejdemyr.github.io/r-tutorials/statistics/tutorial8.html)

Steps	Suggested	for	Conducting a	Propensity Score	Analysis from A Practical	Introduction	Using	R:   

1. Preliminary analysis

Before propensity scores are calculated, it is a good practice to determine if the two groups are balanced. The standardized (mean) difference is a measure of distance between two group means in terms of one or more variables. In practice it is often used as a balance measure of individual covariates before and after propensity score matching. As it is standardized, comparison across variables on different scales is possible. For the standardized difference, absolute scores higher than __25%__ are considered suspect, and may indicate an imbalance for that specific variable. A statistically significant $\chi^2$ will indicate that at least one of the variables included in the model is creating an imbalance between the two groups. 
```{r}
#################################################
# Computing	indices	of co-variate imbalance	before matching 
#################################################
###	1.	Standardized	difference
library(MatchIt)
library(tableone)
data(lalonde)
head(lalonde)
lalonde <- lalonde %>% mutate(
  married = factor(married),
  nodegree = factor(nodegree)
)


## Covariates
vars <- c("age","race","educ","married","nodegree","re74","re75","re78")

## Construct a table
tabUnmatched <- CreateTableOne(vars = vars, strata = "treat", data = lalonde, test = FALSE)
## Show table with SMD
print(tabUnmatched, smd = TRUE)

###	2.	omnibus chi-square	test
library("RItools")
xBalance(treat	~ age	+ educ	+ nodegree	+ re74	+ re75,	data	= lalonde,	report	= c("chisquare.test"))
```
```{r}
#--- Outcome	model	using	Regression	analysis	
reg	<- lm(re78	~ treat	+ age	+ educ	+ nodegree	+ re74	+ re75	+ married	+ race,	data = lalonde)
summary(reg)
```



2. Estimation of Propensity scores
In this step, the propensity score is estimated.
```{r}
### Calculates	the	propensity	score
ps <- glm(treat	~ age	+ educ	+ nodegree	+ re74	+ re75,	data	=lalonde,	family	= binomial())
summary(ps)

### Attach	the	predicted	propensity	score	to	the	data
lalonde$psvalue	<- predict(ps,	type	= "response")
### Back	to	back	histogram
Hmisc::histbackback(split(lalonde$psvalue,	lalonde$treat),	main= "Propensity score before matching",	xlab=c("control",	"treatment"))
```


3. Propensity Score matching
```{r}
#---Match	using	near-neighbor
m.nn	<- matchit(treat	~ age	+ educ	+ nodegree	+ re74	+ re75,	data	=lalonde,	method= "nearest",	ratio	= 1)
summary(m.nn)
plot(m.nn,type="jitter")
```

After the match has been conducted, both techniques-graphical and omnibus test, are used again to determine that all the critical variables have been balanced.
```{r}
match.data	= match.data(m.nn)
#---Computing	indices	of	covariate	imbalance	after	matching
###	1.	Standardized	difference
## Construct a table
tabUnmatched <- CreateTableOne(vars = vars, strata = "treat", data = match.data, test = FALSE)
## Show table with SMD
print(tabUnmatched, smd = TRUE)

###	2.	chi-square	test
xBalance(treat	~ age	+ educ	+ nodegree	+ re74	+ re75,	data	= match.data,	report	= c("chisquare.test"))
```

As can be observed in this figure, although the chisquare test indicates no significance, thus suggesting equivalence between the groups, the standardized difference test shows that there are some variables with a large difference (i.e., race, married) that can still be improved.

```{r}
Hmisc::histbackback(split(match.data$psvalue,	match.data$treat),	main= "Propensity score after matching",	xlab=c("control",	"treatment"))
```
The selection bias has been reduced substantially after matching propensity score.

4. Outcome analysis 

5. Sensitivity analysis 
  

  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
